<!--
vignette: >
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteIndexEntry{TSRchitect}
    %\VignettePackage{TSRchitect}
    %\VignetteIndexExtry{TSRchitect}
    \usepackage[utf8]{inputenc}
-->
```{r setup, include=FALSE}
library(knitr)
```

## TSRchitect Package Vignette

### R. Taylor Raborn, Indiana University

### 13 January 2017

TSRchitect is an R package for analyzing diverse types of high-throughput transcription start site (TSS) profiling datasets. TSRchitect can handle TSS profiling experiments that contain either single-end orb paired-end sequence. In recent years, large-scale TSS profiling data has characterized the landscape of transcription initiation at high resolution, identifying promoter architecture in a number of eukaryotic model systems, including human, mouse, fruit fly and worm.

Examples of TSS profiling data types that TSRchitect is capable of handling are as follows:

- CAGE (Cap Analysis of Gene Expression) [Single-end]
- PEAT (Paired-end Analysis of Transcription) [Paired-end]
- RAMPAGE (RNA Annotation and Mapping of Promoters for Analysis of Gene Expression) [Paired-end]
- TSS-seq [Single-end]

TSRchitect provides the capability to efficiently identify putative promoters---which we call transcription start regions (TSRs)---from TSS profiling experiments. TSRchitect possesses the flexibility to accomodate datasets, including biological replicates and multiple tissues/conditions, that were generated in a variety of model organisms and genome assemblies, requiring only aligned TSS profiling information (in BAM format) as the initial input. To aid the downstream analysis of identified promoters, TSRchitect calcualtes a variety of TSR properties that have been shown to be associated with promoter architecture, including TSR activity, width and the Shape Index (SI). Finally, TSRchitect's output is compatible with popular differential expression analyses such as edgeR, assisting in downstream analysis to identify TSRs that are differentially active in one sample versus another. In addition to this vignette, the TSRchitect Reference Manual is available as part of the package's online documentation.

To install TSRchitect, please type the following from an R console:

```{r eval=FALSE}
library(devtools) #loads devtools (please install this package if you haven't already done so)
install_github("brendelgroup/TSRchitect")
```

## Example 1: Identifying promoters from RAMPAGE data derived from two human cell lines.

RAMPAGE is a TSS profiling method that identifies promoters at large-scale using a cap-based library construction method that is adapted for paired-end sequencing. Developed recently by Batut and Gingeras (2013), it has become a popular method for promoter identification and is currently part of the data compredium in the latest edition of the ENCODE project.

In this example we will process RAMPAGE data derived from two immortalized human cell lines with TSRchitect. The experiments selected for this vignette are part of the ENCODE project and is publically available online at the [ENCODE Experiment matrix](https://www.encodeproject.org/matrix/?type=Experiment). The two samples come from HT1080 cells, which is a well-characterized fibrosarcoma cell line, and NCI-H460 cells, which are derived from a male large cell lung carcinoma. 

Before we begin, please download the RAMPAGE datasets (which were aligned to GRCh38 and are in BAM format) from the following locations:

HT1080:
-[ENCFF474YPI-rep1](https://www.encodeproject.org/files/ENCFF474YPI/@@download/ENCFF474YPI.bam)
-[ENCFF242UWH-rep2](https://www.encodeproject.org/files/ENCFF242UWH/@@download/ENCFF242UWH.bam)

NCI-H460:
-[NCI-H460-rep1](https://www.encodeproject.org/files/ENCFF214GWH/@@download/ENCFF214GWH.bam)
-[NCI-H460-rep2](https://www.encodeproject.org/files/ENCFF265SGZ/@@download/ENCFF265SGZ.bam)

Please download these to your machine and place them in a new folder called ```bam_files/``` inside the demo/ folder.

Now that we have our input files prepared, we can load TSRchitect into our workspace, along with another library required for parallelization:

```{r eval=FALSE}
library(TSRchitect) #loading TSRchitect
library(doParallel) #loading the doParallel package, which is requried for parallel processing
```

To implement parallelization, we'll need to set up a cluster of 6 nodes first. (Please adjust the number of clusters as you desire and your compute permits):

```{r eval=FALSE}
mycl <- makeCluster(6,type="FORK",outfile="")
registerDoParallel(mycl)
```

Next we'll create the dedicated S4 object---called the tssObj---that TSRchitect's functions will be performed on.

```{r eval=FALSE}
initializeExp(expTitle ="Human RAMPAGE", experimentName="Hs_RAMPAGE", expDir="vignette_files/", isPairedEnd=TRUE) 
```

This will create an object called `Hs_RAMPAGE`, in your workspace. We specified the directory "vignette_files" as the location of the bam files to be imported, and have given it the rather general title "Human RAMPAGE". Before we continue, let's take a look look at it to make sure it appears:

```{r eval=FALSE}
ls()
```
Unless you have been doing other work in your R console, that command should just return our tssObj S4 object:

```{r eval=FALSE
[1] "Hs_RAMPAGE"
```
Now we need to provide the sample names and specify which samples are biological replicates. In this case we are working with 4 total datasets and 2 samples in duplicate, and we provide the names as follows:

```{r eval=FALSE}
setSampleID(Hs_RAMPAGE, c("HT1080-1","HT1080-2","H460-1", "H460-2"), c(1,1,2,2))
```

Now that we have completed the above commands, we can proceed with importing the bam files using the function `importBam()`. This will attach GenomicAlignments objects (representing the four bam files in this example) to your tssObject. 

```{r eval=FALSE}
importBam(Hs_RAMPAGE)

... importBam ...

Importing paired-end reads ...

TSS data were specified to be paired-end read alignments.
Beginning import of  4  bam files ...
```

This step will take between ten and fifteen minutes, so feel free to take a break and resume when the job is complete. You will receive a message on your console when all four files have been imported. 

```{r eval=FALSE}
Done. Alignment data from  4  bam files have been attached to tssObject
object " Hs_RAMPAGE ".
 Done.
-------------------------------------------------------------------------------- 
```

Now that the alignment files have been imported and attached to our tssObj S4 object, we continue by computing the TSSs from the bam files. This will take approximately 5 minutes. 

```{r eval=FALSE}
bamToTSS(experimentName=Hs_RAMPAGE)
}
```

Next we will calculate the abundance of each tag in our TSS datasets.

```{r eval=FALSE}
processTSS(experimentName=Hs_RAMPAGE, parallel=TRUE, tssSet="all", writeTable=TRUE)
```

Since we specified 'writeTable=TRUE', files containing TSS abundance will be written into your working directory. 

Once we have calculated the abundance of each TSS we can calculate TSRs (promoters) on each of the 4 separate datasets. We have selected a tagCount threshold of 25 tags in order for a TSS to be considered.

```{r eval=FALSE}
determineTSR(experimentName = Hs_RAMPAGE, parallel = TRUE, tsrSetType = "replicates", tssSet="all", tagCountThreshold=25, clustDist=20, writeTable=TRUE)
```

Now we need to merge the replicates into samples, as we have specified on your tssObj S4 object.

```{r eval=FALSE}
mergeSampleData(experimentName=Hs_RAMPAGE)
```

```{r eval=FALSE}
determineTSR(experimentName = Hs_RAMPAGE, parallel = TRUE, tsrSetType = "merged", tssSet="all", tagCountThreshold=40, clustDist=20, writeTable=TRUE)
```

```{r eval=FALSE}
addTagCountsToTSR(experimentName=Hs_RAMPAGE, tsrSetType="merged", tsrSet=3, tagCountThreshold=40, writeTable=TRUE)
```



You can use your own CSS file instead of the built-in style in the **markdown** package -- just set the option `markdown.HTML.stylesheet`, e.g.

```{r css, eval=FALSE}
options(markdown.HTML.stylesheet = 'path/to/a/custom/style.css')
```
