<!--
vignette: >
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteIndexEntry{TSRchitect}
    %\VignettePackage{TSRchitect}
    %\VignetteIndexExtry{TSRchitect}
    \usepackage[utf8]{inputenc}
-->
```{r setup, include=FALSE}
library(knitr)
```

## TSRchitect Package Vignette

### R. Taylor Raborn, Indiana University

### 13 January 2017

TSRchitect is an R package for analyzing diverse types of high-throughput transcription start site (TSS) profiling datasets. TSRchitect can handle TSS profiling experiments that contain either single-end orb paired-end sequence. In recent years, large-scale TSS profiling data has characterized the landscape of transcription initiation at high resolution, identifying promoter architecture in a number of eukaryotic model systems, including human, mouse, fruit fly and worm.

Examples of TSS profiling data types that TSRchitect is capable of handling are as follows:

- CAGE (Cap Analysis of Gene Expression)
- PEAT (Paired-end Analysis of Transcription)
- RAMPAGE (RNA Annotation and Mapping of Promoters for Analysis of Gene Expression)
- TSS-seq 

TSRchitect provides the capability to efficiently identify putative promoters---which we call transcription start regions (TSRs)---from TSS profiling experiments. TSRchitect possesses the flexibility to accomodate datasets, including biological replicates and multiple tissues/conditions, that were generated in a variety of model organisms and genome assemblies, requiring only aligned TSS profiling information (in BAM format) as the initial input. To aid the downstream analysis of identified promoters, TSRchitect calcualtes a variety of TSR properties that have been shown to be associated with promoter architecture, including TSR activity, width and the Shape Index (SI). Finally, TSRchitect's output is compatible with popular differential expression analyses such as edgeR, assisting in downstream analysis to identify TSRs that are differentially active in one sample versus another. In addition to this vignette, the TSRchitect Reference Manual is available as part of the package's online documentation.

To install TSRchitect, you can type the following from an R console:

```{r eval=FALSE}
library(devtools) #loads devtools (please install this package if you haven't already done so)
install_github("brendelgroup/TSRchitect")
```

## Examples

Below are some code chunks as examples.

```{r hello, results='asis'}
if (TRUE) cat('_hello_ **markdown**!', '\n')
```

Normally you do not need any chunk options.

```{r test, collapse=TRUE}
1+1
10:1
rnorm(5)^2
strsplit('hello, markdown vignettes', '')
```

Feel free to draw beautiful plots and write math $P(X>x)=\alpha/2$.

```{r}
n=300; set.seed(123)
par(mar=c(4,4,.1,.1))
plot(rnorm(n), rnorm(n), pch=21, cex=5*runif(n), col='white', bg='gray')
```

You can use your own CSS file instead of the built-in style in the **markdown** package -- just set the option `markdown.HTML.stylesheet`, e.g.

```{r css, eval=FALSE}
options(markdown.HTML.stylesheet = 'path/to/a/custom/style.css')
```
