---
title: "TSRchitect Introduction "
author: "R. Taylor Raborn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TSRchitect vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## R. Taylor Raborn and Volker P. Brendel
## Department of Biology, Indiana University

### January 25, 2017

TSRchitect is an R package for analyzing diverse types of high-throughput transcription start site (TSS) profiling datasets.
TSRchitect can handle TSS profiling experiments that contain either single-end or paired-end sequence reads, such as CAGE, RAMPAGE, PEAT and others. TSRchitect is an open-source bioinformatics package that is intended to support large-scale, reproducible analysis of TSS profiling data in a broad array of eukaryotic model systems.

The TSRchitect User's Guide is available; it goes through several well-documented examples of TSRchitect.

To open the TSRchitect User's Guide on your machine, enter the following:

```{r eval=TRUE}
library(TSRchitect)
TSRchitectUsersGuide()
```

## Getting started

```{r eval=TRUE}
library(TSRchitect)
```

Now that you have loaded TSRchitect, we can proceed with a small example.

