<!--
vignette: >
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteIndexEntry{TSRchitect}
    %\VignettePackage{TSRchitect}
    %\VignetteIndexExtry{TSRchitect}
    \usepackage[utf8]{inputenc}
-->
```{r setup, include=FALSE}
library(knitr)
```

This will be the Vignette for TSRchitect. What follows is a template file from the following [site](https://raw.githubusercontent.com/yihui/knitr/master/vignettes/knitr-markdown.Rmd), and will be filled in with the TSRchitect vignette material shortly.

## Package vignettes

To enable Markdown vignettes in an R package, you need to

- add `*.Rmd` files under the `vignettes` directory
- add `VignetteBuilder: knitr` to the `DESCRIPTION` file
- specify the vignette engine `\VignetteEngine{knitr::rmarkdown}` in the `Rmd` files (inside HTML comments)

## View vignettes

And R will load the **knitr** package to build these vignettes to HTML files, and you can see them when you open the HTML help:

```{r eval=FALSE}
help(package = 'YourPackage', help_type = 'html')
# or see a standalone list of vignettes
browseVignettes('YourPackage')
```

## Examples

Below are some code chunks as examples.

```{r hello, results='asis'}
if (TRUE) cat('_hello_ **markdown**!', '\n')
```

Normally you do not need any chunk options.

```{r test, collapse=TRUE}
1+1
10:1
rnorm(5)^2
strsplit('hello, markdown vignettes', '')
```

Feel free to draw beautiful plots and write math $P(X>x)=\alpha/2$.

```{r}
n=300; set.seed(123)
par(mar=c(4,4,.1,.1))
plot(rnorm(n), rnorm(n), pch=21, cex=5*runif(n), col='white', bg='gray')
```

You can use your own CSS file instead of the built-in style in the **markdown** package -- just set the option `markdown.HTML.stylesheet`, e.g.

```{r css, eval=FALSE}
options(markdown.HTML.stylesheet = 'path/to/a/custom/style.css')
```
